---
title: "COVID19_Johns_Hopkins"
author: "Carlos Barron"
date: "2022-11-08"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
webshot::install_phantomjs()
library(tidyverse)
library(knitr)
library(webshot)
library(lubridate)
library(plotly)
library(dplyr)
library(ggplot2)
library(scales)
```

## Introduction
After what was analyzed in the lectures, I’m interested to see if we can visualize a high level of comorbidity between smoking, exercising and COVID-19 within the dataset. Before I try to put a model in place, I need to find external datasets about smoking and physical exercise such that I may be able to make any further analysis. I will attempt to relate the datasets through the country column.

## Data
**COVID-19**

The COVID-19 data for this report consists of 2 CSVs that you can find [here.](https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series) 

Each one represents the confirmed cases and deaths worldwide.
```{r load_covid19_data, include=FALSE}
urls = str_c(
  "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/",    c(
    "time_series_covid19_confirmed_global.csv",
    "time_series_covid19_deaths_global.csv"
  )
)
confirmed_global = read_csv(urls[1])
deaths_global = read_csv(urls[2])


confirmed_global <- confirmed_global %>% pivot_longer(cols = -c(`Province/State`,`Country/Region`, Lat, Long), names_to = "date", values_to = "cases") %>% rename("country"="Country/Region", "province"="Province/State")  %>%  mutate(
  date = as.Date(date, "%m/%d/%y"),
  year = format(date, format="%Y")
) %>% group_by(country, date) %>% summarise(country = last(country), date = last(date), year = last(year), cases = sum(cases)) %>% group_by(country, year) %>% summarise(country = last(country), year = last(year), cases = last(cases))

deaths_global <- deaths_global %>% pivot_longer(cols = -c(`Province/State`,`Country/Region`, Lat, Long), names_to = "date", values_to = "deaths")%>% rename("country"="Country/Region", "province"="Province/State")  %>%  mutate(
  date = as.Date(date, "%m/%d/%y"),
  year = format(date, format="%Y")
) %>% group_by(country, date) %>% summarise(country = last(country), date = last(date), year = last(year), deaths = sum(deaths)) %>% group_by(country, year) %>% summarise(country = last(country), year = last(year), deaths = last(deaths))

countries = c('Peru','Mexico','China', 'Hungary', 'Romania', 'South',' Africa','Italy','Brazil','Poland','Colombia', 'Germany', 'Chile', 'Argentina', 'Belgium','Canada','Spain', 'France', 'Saudi Arabia', 'Australia', 'India', 'Sweden', 'Japan', 'Malaysia', 'Netherlands')
```

Confirmed Cases
```{r confirmed_cases, include=TRUE}
confirmed_global
```
Confirmed Deaths
```{r confirmed_deaths, include=TRUE}
confirmed_global
```


**[Tobacco Atlas](https://tobaccoatlas.org/about/)**

For cigarette consumption I will use the dataset avaialable throught theTobacco Atlas available [here.](https://tobaccoatlas.org/wp-content/uploads/2022/05/data-z9rR9.csv)

Fields

1. Country
2. Average daily number of cigarettes consumed per adult (15+ yr) smoker, 2019

```{r cigar_consumption_data_load, include=FALSE}
avg_daily_cigar_consumption = read_csv("https://tobaccoatlas.org/wp-content/uploads/2022/05/data-z9rR9.csv") %>% rename("country"="Country", "consumption" = "Cigarette Consumption")
```

```{r cigar_consumption, include=TRUE}
avg_daily_cigar_consumption = avg_daily_cigar_consumption %>% filter(avg_daily_cigar_consumption$country %in% countries) %>% select(c(country, `consumption`))

fig <- plot_ly(
  avg_daily_cigar_consumption,
  x = ~country, 
  y = ~consumption,
  name = "Average Daily Cigar Consumption by Country",
  type = "bar",
  orientation="v"
) %>% layout(xaxis = list(categoryorder = "total descending"))
fig
```

**[Ipsos Global Advisor](https://www.ipsos.com/en/about-us)**

Global Views on Exercise and Team Sports

For the exercise information I will use the dataset available [here.](https://www.ipsos.com/sites/default/files/ct/news/documents/2021-08/Global-Views-on-Sports-and-Exercise-Ipsos.pdf)

Fields

1. Country
2. Mean Number of Hours Physical Excercise Per Week

```{r excercise_data_load, include=FALSE}
weekly_excercise = read_csv("./data/ipsos_global_views_exercise.csv")
```

```{r weekly_excercise, include=TRUE}
weekly_excercise = weekly_excercise %>% filter(weekly_excercise$country %in% countries) %>% select(c(country, weekly_excercise_hours_mean))

fig <- plot_ly(
  weekly_excercise,
  x = ~country, 
  y = ~weekly_excercise_hours_mean,
  name = "Weekly Excercise Hours Mean by Country",
  type = "bar",
  orientation="v"
) %>% layout(xaxis = list(categoryorder = "total descending"))
fig
```

**[United Nations](https://www.un.org/en/desa/about-us)**

Department of Economic and Social Affairs, World Population Prospects 2022

For the age information I will use the dataset available [here.](https://population.un.org/wpp/Download/Standard/CSV/)

Fields

1. Country
2. Median Age

```{r median_age_data_load, include=FALSE}
median_age_by_country = read_csv("./data/median_age_country.csv")
```

```{r median_age, include=TRUE}
median_age_by_country = median_age_by_country %>% filter(median_age_by_country$year == 2020) %>% select(-c(year))
median_age_by_country = median_age_by_country %>% filter(median_age_by_country$country %in% countries) %>% select(c(country, median_age))

fig <- plot_ly(
  median_age_by_country,
  x = ~country, 
  y = ~median_age,
  name = "Median Age by Country",
  type = "bar",
  orientation="v"
) %>% layout(xaxis = list(categoryorder = "total descending"))
fig
```

## Table Joins
Now that we have the data loaded, let's join all of the different tables by country and year
```{r data_merge, include=TRUE}
covid_stats = inner_join(
  confirmed_global, deaths_global, by = c('country','year')
) %>%  mutate(mortality = (
  deaths * 100) / cases
)

covid_stats = inner_join(
  covid_stats, avg_daily_cigar_consumption, by = c('country')
)
covid_stats = inner_join(
  covid_stats, weekly_excercise, by = c('country')
)
covid_stats = inner_join(
  covid_stats, median_age_by_country, by = c('country')
)
covid_stats
```
## Mortality
Let's graph the mortality rate before the Vaccine came out (August 2021).
```{r mortality, include=TRUE}
covid_stats_2020 <- filter(covid_stats, year == 2020)
fig <- plot_ly(
  covid_stats,
  x = ~country, 
  y = ~mortality,
  name = "Mortality by Country",
  type = "bar",
  orientation="v"
) %>% layout(xaxis = list(categoryorder = "total descending"))
fig
```


## Models
*This is the model summary for cigarette consumption*
```{r model_consumption, include=TRUE}
covid_model_consumption <- lm(deaths ~ consumption, data = covid_stats)
summary(covid_model_consumption)
```
*This is the model summary for weekly excercise mean*
```{r model_excercise, include=TRUE}
covid_model_excercise <- lm(deaths ~ weekly_excercise_hours_mean, data = covid_stats)
summary(covid_model_excercise)
```
*And this is the model summary for the median age*
```{r model_age, include=TRUE}
covid_model_age <- lm(deaths ~ median_age, data = covid_stats)
summary(covid_model_age)
```

```{r prediction_consumption, include=TRUE}
global_total_deaths_w_pred <- covid_stats %>% mutate(
  consumption_prediction = 194815.29 -(113.89 * consumption),
  excercise_prediction = 162385 -(10296 * weekly_excercise_hours_mean),
  age_prediction = 402604 -(8124 * median_age)
)
```

## Graphing the models

*Cigarette Consumption*

```{r chart_consumption_prediction, include=TRUE}
global_total_deaths_w_pred %>% ggplot() + geom_point(
  aes(x=consumption, y=deaths),
  color = "blue"
) + geom_point(
  aes(x=consumption, y=consumption_prediction),
  color = "red"
) + scale_y_continuous(labels = label_number(suffix = "K", scale = 1e-3))
```

*Weekly Excercise*

```{r chart_excercise_prediction, include=TRUE}
global_total_deaths_w_pred %>% ggplot() + geom_point(
  aes(x=weekly_excercise_hours_mean, y=deaths),
  color = "blue"
) + geom_point(
  aes(x=weekly_excercise_hours_mean, y=excercise_prediction),
  color = "red"
) + scale_y_continuous(labels = label_number(suffix = "K", scale = 1e-3))
```

*Age*

```{r chart_age_prediction, include=TRUE}
global_total_deaths_w_pred %>% ggplot() + geom_point(
  aes(x=median_age, y=deaths),
  color = "blue"
) + geom_point(
  aes(x=median_age, y=age_prediction),
  color = "red"
) + scale_y_continuous(labels = label_number(suffix = "K", scale = 1e-3))
```

## Bias & Conclusion
- The cigarette consumption and age are significant but my model had low R-squared values.
- The excercise doesn't look to have a significance and had a low R-squared value.
- There are obviously other very important factors to be considered such as public health governance, and the timeline of infections for each country during the pandemic

Finally, please find the session info below.
```{r session_info, include=TRUE}
sessionInfo()
```