---
title: "COVID-19 Johns Hopkins Data Analysis"
author: "Carlos Barron"
date: "2022-11-16"
output:
  html_document:
  pdf_document: default
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(webshot)
library(knitr)
library(lubridate)
library(dplyr)
library(ggplot2)
library(scales)
library(rmarkdown)
```

## Introduction
After what was analyzed in the lectures, I’m interested to see if we can visualize a high level of comorbidity between smoking, exercising and COVID-19 within the dataset. Before I try to put a model in place, I need to find external datasets about smoking and physical exercise such that I may be able to make any further analysis. I will attempt to relate the datasets through the country column.

## Data
**COVID-19**

The COVID-19 data for this report consists of 2 CSVs that you can find [here.](https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series) 

Each one represents the confirmed cases and deaths worldwide.
```{r load_covid19_data, include=FALSE}
urls = str_c(
  "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/",    c(
    "time_series_covid19_confirmed_global.csv",
    "time_series_covid19_deaths_global.csv"
  )
)
confirmed_global = read_csv(urls[1])
deaths_global = read_csv(urls[2])


confirmed_global <- confirmed_global %>% pivot_longer(cols = -c(`Province/State`,`Country/Region`, Lat, Long), names_to = "date", values_to = "cases") %>% rename("country"="Country/Region", "province"="Province/State")  %>%  mutate(
  date = as.Date(date, "%m/%d/%y"),
  year = format(date, format="%Y")
) %>% group_by(country, date) %>% summarise(country = last(country), date = last(date), year = last(year), cases = sum(cases)) %>% group_by(country, year) %>% summarise(country = last(country), year = last(year), cases = last(cases)) %>% mutate(
  cases_K = cases * 1e-3,
  cases_M = cases * 1e-6
)

deaths_global <- deaths_global %>% pivot_longer(cols = -c(`Province/State`,`Country/Region`, Lat, Long), names_to = "date", values_to = "deaths")%>% rename("country"="Country/Region", "province"="Province/State")  %>%  mutate(
  date = as.Date(date, "%m/%d/%y"),
  year = format(date, format="%Y")
) %>% group_by(country, date) %>% summarise(country = last(country), date = last(date), year = last(year), deaths = sum(deaths)) %>% group_by(country, year) %>% summarise(country = last(country), year = last(year), deaths = last(deaths)) %>% mutate(
  deaths_K = deaths * 1e-3,
  deaths_M = deaths * 1e-6
)

countries = c('Peru','Mexico','China', 'Hungary', 'Romania', 'South',' Africa','Italy','Brazil','Poland','Colombia', 'Germany', 'Chile', 'Argentina', 'Belgium','Canada','Spain', 'France', 'Saudi Arabia', 'Australia', 'India', 'Sweden', 'Japan', 'Malaysia', 'Netherlands')
```


Confirmed Cases
```{r eval=knitr::is_html_output(), include=TRUE}
paged_table(confirmed_global, options = list(rows.print = 15, cols.print = 10))
```

```{r eval=knitr::is_latex_output(), include=TRUE}
confirmed_global
```

Confirmed Deaths
```{r eval=knitr::is_html_output(), include=TRUE}
paged_table(deaths_global, options = list(rows.print = 15, cols.print = 10))
```

```{r eval=knitr::is_latex_output(), include=TRUE}
deaths_global
```

```{r data_loads_chart_creation, include=FALSE}
avg_daily_cigar_consumption = read_csv("https://tobaccoatlas.org/wp-content/uploads/2022/05/data-z9rR9.csv") %>% rename("country"="Country", "consumption" = "Cigarette Consumption")

avg_daily_cigar_consumption = avg_daily_cigar_consumption %>% filter(avg_daily_cigar_consumption$country %in% countries) %>% select(c(country, `consumption`))

avg_daily_cigar_chart = ggplot(
  avg_daily_cigar_consumption, 
  aes(x=reorder(country, -consumption), y=consumption)
) + geom_bar(stat = "identity") + theme(axis.text.x = element_text(angle = 90, hjust=1)) + labs(title = "Cigarette Consumption by Country", x = "Country", y = "Cigarettes Consumed")

weekly_excercise = read_csv("./data/ipsos_global_views_exercise.csv")
weekly_excercise = weekly_excercise %>% filter(weekly_excercise$country %in% countries) %>% select(c(country, weekly_excercise_hours_mean))

weekly_excercise_chart = ggplot(
  weekly_excercise, 
  aes(x=reorder(country, -weekly_excercise_hours_mean), y=weekly_excercise_hours_mean)
) + geom_bar(stat = "identity") + theme(axis.text.x = element_text(angle = 90, hjust=1)) + labs(title = "Weekly Excercise Hours Mean by Country", x = "Country", y = "Weekly Excercise Hours")

median_age_by_country = read_csv("./data/median_age_country.csv")
median_age_by_country = median_age_by_country %>% filter(median_age_by_country$year == 2020) %>% select(-c(year))
median_age_by_country = median_age_by_country %>% filter(median_age_by_country$country %in% countries) %>% select(c(country, median_age))

median_age_chart = ggplot(
  median_age_by_country, 
  aes(x=reorder(country, -median_age), y=median_age)
) + geom_bar(stat = "identity") + theme(axis.text.x = element_text(angle = 90, hjust=1)) + labs(title = "Median Age by Country", x = "Country", y = "Median Age")
```

### [Tobacco Atlas](https://tobaccoatlas.org/about/)

For cigarette consumption I will use the dataset avaialable throught theTobacco Atlas available [here.](https://tobaccoatlas.org/wp-content/uploads/2022/05/data-z9rR9.csv)

Fields

1. Country
2. Average daily number of cigarettes consumed per adult (15+ yr) smoker, 2019

```{r cigar_consumption_chart, include=TRUE}
avg_daily_cigar_chart
```

### [Ipsos Global Advisor](https://www.ipsos.com/en/about-us)

Global Views on Exercise and Team Sports

For the exercise information I will use the dataset available [here.](https://www.ipsos.com/sites/default/files/ct/news/documents/2021-08/Global-Views-on-Sports-and-Exercise-Ipsos.pdf)

Fields

1. Country
2. Mean Number of Hours Physical Excercise Per Week

```{r weekly_excercise_chart, include=TRUE}
weekly_excercise_chart
```

### [United Nations](https://www.un.org/en/desa/about-us)

Department of Economic and Social Affairs, World Population Prospects 2022

For the age information I will use the dataset available [here.](https://population.un.org/wpp/Download/Standard/CSV/)

Fields

1. Country
2. Median Age

```{r median_age_chart, include=TRUE}
median_age_chart
```

## Table Joins
Now that we have the data loaded, let's join all of the different tables by country and year
```{r data_merge_load, include=FALSE}
covid_stats = inner_join(
  confirmed_global, deaths_global, by = c('country','year')
) %>%  mutate(mortality_K = (
  deaths * 1000) / cases
)

covid_stats = inner_join(
  covid_stats, avg_daily_cigar_consumption, by = c('country')
)
covid_stats = inner_join(
  covid_stats, weekly_excercise, by = c('country')
)
covid_stats = inner_join(
  covid_stats, median_age_by_country, by = c('country')
)
```


```{r eval=knitr::is_html_output(), include=TRUE}
paged_table(covid_stats, options = list(rows.print = 15, cols.print = 10))
```

```{r eval=knitr::is_latex_output(), include=TRUE}
paged_table(covid_stats, options = list(rows.print = 15, cols.print = 10))
```
## Mortality
Let's graph the mortality rate before the Vaccine came out (August 2021).
```{r mortality_load, include=FALSE}
covid_stats_2020 <- filter(covid_stats, year == 2020)
covid_stats_2021 <- filter(covid_stats, year == 2021)
covid_stats_2022 <- filter(covid_stats, year == 2022)
mortality_chart = ggplot(
  covid_stats_2020, 
  aes(x=reorder(country, -mortality_K), y=mortality_K)
) + geom_bar(stat = "identity") + theme(axis.text.x = element_text(angle = 90, hjust=1)) + labs(title = "Mortality by Country 2020", x = "Country", y = "Mortality %")
```

```{r mortality_load_chart, include=TRUE}
mortality_chart
```


## Models & Analysis
Let's start by making an analysis on how much mortality and cigarette consumption is related.
```{r analysis_chart_loads, include=FALSE}
consumption_analysis_load = covid_stats_2022 %>% ggplot() + geom_point(
  aes(x=consumption, y=mortality_K),
  color = "blue"
) + geom_smooth(
  aes(x=consumption, y=mortality_K),
  color = "red",
  method = "lm"
)
excercise_chart_load = covid_stats_2022 %>% ggplot() + geom_point(
  aes(x=weekly_excercise_hours_mean, y=mortality_K),
  color = "blue"
) + geom_smooth(
  aes(x=weekly_excercise_hours_mean, y=mortality_K),
  color = "red",
  method = "lm"
)
age_chart_load = covid_stats_2022 %>% ggplot() + geom_point(
  aes(x=median_age, y=mortality_K),
  color = "blue"
) + geom_smooth(
  aes(x=median_age, y=mortality_K),
  color = "red",
  method = "lm"
)
```

```{r analysis_model_loads, include=FALSE}
covid_model_consumption <- lm(mortality_K ~ consumption, data = covid_stats_2022)
covid_model_excercise <- lm(mortality_K ~ weekly_excercise_hours_mean, data = covid_stats_2022)
covid_model_age <- lm(mortality_K ~ median_age, data = covid_stats_2022)
```

```{r consumption_analysis_chart, include=TRUE}
consumption_analysis_load
```
*This is the model summary for cigarette consumption*
```{r model_consumption, include=TRUE}
summary(covid_model_consumption)
```
Now let's see how mortality and physical exercise is related.
```{r excercise_chart_display, include=TRUE}
excercise_chart_load
```

*This is the model summary for weekly excercise mean*
```{r model_excercise, include=TRUE}
summary(covid_model_excercise)
```
Finally let's see how mortality and age is related.
```{r excercise_analysis, include=TRUE}
age_chart_load
```

*And this is the model summary for the median age*
```{r model_age, include=TRUE}
summary(covid_model_age)
```
Let's build the prediction into our dataset.
```{r prediction_consumption, include=TRUE}
global_total_deaths_w_pred <- covid_stats_2022 %>% mutate(
  consumption_prediction = 21.789989 -(0.0009448 * consumption),
  excercise_prediction = 18.3313 -(0.7268 * weekly_excercise_hours_mean),
  age_prediction = 52.8234 -(1.0447 * median_age)
)
```

## Graphing the models
```{r model_charts, include=FALSE}
cig_model_chart = global_total_deaths_w_pred %>% ggplot() + geom_point(
  aes(x=consumption, y=mortality_K),
  color = "blue"
) + geom_point(
  aes(x=consumption, y=consumption_prediction),
  color = "red"
)
excercise_model_chart = global_total_deaths_w_pred %>% ggplot() + geom_point(
  aes(x=weekly_excercise_hours_mean, y=mortality_K),
  color = "blue"
) + geom_point(
  aes(x=weekly_excercise_hours_mean, y=excercise_prediction),
  color = "red"
)
age_model_chart = global_total_deaths_w_pred %>% ggplot() + geom_point(
  aes(x=median_age, y=mortality_K),
  color = "blue"
) + geom_point(
  aes(x=median_age, y=age_prediction),
  color = "red"
)
```

*Cigarette Consumption*
```{r chart_consumption_prediction, include=TRUE}
cig_model_chart
```

*Weekly Excercise*
```{r chart_excercise_prediction, include=TRUE}
excercise_model_chart
```

*Age*
```{r chart_age_prediction, include=TRUE}
age_model_chart
```

## Bias & Conclusion
- After these iterations, the models are not statistically significant and had low R-squared values.
- For next iterations I would replace some of the data obtained (exercise and cigarette consumption)
- Cigarette consumption by adult might not be a great fit since it does not give a clear indication of the percentage of smokers by country.
- Getting information for the percentage of population who excercises might be tricky.
- An important factor to take into account is the timeline on which the vaccines started to roll out by country, this could have introduced noise in my implementation.
- From the list of countries in the implementation, all had different isolation strategies that should also be considered.
- The quality and saturation of the public health care system is another important factor to take into account.

Finally, please find the session info below.
```{r session_info, include=TRUE}
sessionInfo()
```